CytoGAN: Generative Modeling of Cell Images

### 2017-11-23 paper/paper.tex
Updated paper

% to compile a camera-ready version, add the [final] option, e.g.:
% \usepackage[final]{nips_2017}
\usepackage{nips_2017}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsmath}
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{todonotes}

@@

%\PassOptionsToPackage{numbers}{natbib}
%\usepackage[numbers]{natbib}
%\usepackage[natbib, maxcitenames=3, style=numeric]{biblatex}
%\setcitestyle{numbers}
%\bibliographystyle{abbrvnat}

\title{CytoGAN: Generative Modeling of Cell Images}
% * <anne@broadinstitute.org> 2017-10-15T19:01:17.646Z:
%
% > CytoGAN
% Is it necessary to give it a name? This seems overly common to me, but then again maybe it's the convention and you just go with it.
%
% ^.

\author{
  Peter Goldsborough\thanks{This work has been conducted as part of an internship at the Imaging Platform of the Broad Institute.}\\
  %Imaging Platform \\
  Broad Institute of MIT and Harvard \\
  %Cambridge, MA, USA \\
  \texttt{pgoldsbo@broadinstitute.org} \\
  \And
  Nick Pawlowski \\
  %Department of Computing \\
  Imperial College London \\
  %London, UK \\
  \texttt{n.pawlowski16@imperial.ac.uk} \\
  \AND
  Juan C Caicedo \& Shantanu Singh \& Anne E Carpenter\\
  %Imaging Platform \\
  Broad Institute of MIT and Harvard \\
  %Cambridge, MA, USA \\
  \texttt{\{jccaicedo, shsingh, anne\}@broadinstitute.org} \\
}

@@

\section{Using GANs for Representation Learning}
Goodfellow et al. \cite{goodfellow2014generative} introduced GANs as a game of
two players: a \emph{generator} $G$ and a \emph{discriminator} or \emph{critic} $D$. The former receives samples $\mathbf{z}$ drawn from a noise prior $P_{\text{noise}}$ which it maps to values $G(\mathbf{z})$ that should resemble elements of some data distribution $P_{\text{data}}$. The discriminator must learn to distinguish such synthetic samples from real values $\mathbf{x} \sim P_{\text{data}}$. The critic's confidence in the realism of the generator's productions is used as feedback to $G$, guiding it to synthesize ever more realistic replicates of samples from the data prior. This procedure is formalized in a zero-sum game,
\vspace{-1pt}
$$\min_G \max_D V(G, D) = \mathbb{E}_{\mathbf{x} \sim P_{\text{data}}}[\log(D(\mathbf{x})] + \mathbb{E}_{\mathbf{z} \sim P_{\text{noise}}}[\log(1 - D(G(\mathbf{z}))].$$
%\vspace{-1pt}
Radford et al. \cite{radford2015unsupervised} first specialized GANs to image synthesis by introducing Deep Convolutional GANs (DCGANs). DCGANs implement the generator and discriminator based on convolutional operations. Derivations of DCGANs such as Least Squares GAN (LSGAN) \cite{mao2016least} or Wasserstein GAN (WGAN) \cite{arjovsky2017wasserstein} tackle instabilities in the training procedure of early generative adversarial models. In our experiments, we found LSGAN to be most stable, in part leading to higher quality generated images than both DCGAN and WGAN.

The original GAN framework does not include an explicit means of performing inference. As such, we require extensions that allow mapping of a sample $\mathbf{x}$ drawn from the data prior to a latent representation via some encoder transformation $E(\mathbf{x})$. A common approach is to interpret the penultimate layer of the discriminator as a latent space. The activations of this layer serve as a source of representation vectors. These latent codes are thought to be meaningful because the discriminator must develop a strong internal representation of its input to succeed at its discrimination task. Furthermore, this method imposes no computational overhead compared to vanilla GANs. More advanced encoders could tie this encoding to a corresponding noise vector.

@@

% Current approaches to morphological profiling extract features that are challenging to translate into biological meaning, such as Zernike moments. While other features capture readily understandable concepts such as the area occupied by the nucleus, even these are difficult to interpret when a given class of cells is defined by several such features in combination. For transfer learning it is nearly impossible to visualize the concept of a feature. GANs present a solution as they hold a generative model that enables the visualization of the impact of perturbations along a feature of the noise space. Further, more advanced GANs integrate an inference method to make the feature and noise space identical such that visualization of feature vectors becomes possible.
%this issue by ,  we found that GAN representations are not yet competitive with those extracted by other approaches, we believe a major benefit provided by generative adversarial models is their interpretability. We showcase this in the following.

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
  	\draw (0, 0) node {\includegraphics[width=0.6\textwidth]{generated}};

@@

\section{Representation Learning for Morphological Profiling}\label{moa}
We test the quality of representations extracted via the discriminator by evaluating their ability to cluster treatments of similar function, here the mechanism-of-action. We obtain treatment profiles by averaging the extracted single cell representations found as intermediate activations of the discriminator. Further, we follow the experimental protocol of \cite{Ljosa2013} and report an average MOA classification accuracy of a leave-one-compound-out cross-validation using a one nearest neighbor classifier.
\begin{table}
  \centering
  \begin{tabular}{lllllll}
  	\toprule
    DCGAN & WGAN & LSGAN & VAE \cite{pawlowski2016msc} & CP \cite{Singh2014} & Deep Transfer \cite{pawlowski2016automating}  & Deep Transfer \cite{ando2017improving} \\
    \midrule
    43 \%& 45 \% & 63 \% & 49 \% & 90 \% & 91 \% & 96 \% \\
    \bottomrule \\
  \end{tabular}
  \caption{Comparison of the mechanism-of-action classification accuracy. CP refers to the results by Singh et al. \cite{Singh2014} using CellProfiler features. The Deep Transfer methods correspond to Deep Feature Transfer as proposed by Pawlowski et al. \cite{pawlowski2016automating} and refined by Ando et al. \cite{ando2017improving}.}
  \label{moa-results}
  \vspace{-0.5cm}
\end{table}

@@

This work covers only a small fraction of possible applications of GANs to the domain of microscopy images. For example, we hope that future work investigates BiGANs\cite{donahue2016adversarial} or other novel solutions to infer latent features with GANs. We believe that improved inference, combined with the interpretable nature of the GAN framework, may enable simulated experiments by performing algebra with vectors corresponding to cell lines, diseases or other perturbations. Finally, we note that deep learning generally flourishes with larger amounts of data. Thus, we conjecture that the performance of the outlined approaches will improve with larger datasets.
% forcing new page so that references are always on Page 5.
\newpage

%\section*{Acknowledgements}
%NP is supported by Microsoft Research through its PhD Scholarship Programme and the EPSRC Centre for Doctoral Training in High Performance Embedded and Distributed Systems (HiPEDS, Grant Reference EP/L016796/1).

>>>>

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final]{nips_2017}
% \usepackage{nips_2017}

\usepackage[dvipsnames]{xcolor}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsmath}
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{todonotes}

\usepackage{tikz}
\usetikzlibrary{3d}
\usetikzlibrary{shapes}
\usepgflibrary{fpu}

@@

% \PassOptionsToPackage{numbers}{natbib}
% \usepackage[numbers]{natbib}
% \usepackage[natbib, maxcitenames=3, style=numeric]{biblatex}
% \setcitestyle{numbers}
% \bibliographystyle{abbrvnat}

\title{CytoGAN: Generative Modeling of Cell Images}

\author{
  Peter Goldsborough\thanks{This work has been conducted as part of an internship at the Imaging Platform of the Broad Institute.}\\
  Imaging Platform \\
  Broad Institute of MIT and Harvard \\
  Cambridge, MA, USA \\
  \texttt{pgoldsbo@broadinstitute.org} \\
  \And
  Nick Pawlowski \\
  Department of Computing \\
  Imperial College London \\
  London, UK \\
  \texttt{n.pawlowski16@imperial.ac.uk} \\
  \AND
  Juan C Caicedo \& Shantanu Singh \& Anne E Carpenter\\
  Imaging Platform \\
  Broad Institute of MIT and Harvard \\
  Cambridge, MA, USA \\
  \texttt{\{jccaicedo, shsingh, anne\}@broadinstitute.org} \\
}

@@

\section{Using GANs for Representation Learning}
Goodfellow et al. \cite{goodfellow2014generative} introduced GANs as a game of
two players: a \emph{generator} $G$ and a \emph{discriminator} or \emph{critic} $D$. The former receives samples $\mathbf{z}$ drawn from a noise prior $P_{\text{noise}}$ which it maps to values $G(\mathbf{z})$ that should resemble elements of some data distribution $P_{\text{data}}$. The discriminator must learn to distinguish such synthetic samples from real values $\mathbf{x} \sim P_{\text{data}}$. The critic's confidence in the realism of the generator's productions is used as feedback to $G$, guiding it to synthesize ever more realistic replicates of samples from the data prior. This procedure is formalized in a zero-sum game,
\vspace{-1pt}
$$\min_G \max_D V(G, D) = \mathbb{E}_{\mathbf{x} \sim P_{\text{data}}}[\log(D(\mathbf{x})] + \mathbb{E}_{\mathbf{z} \sim P_{\text{noise}}}[\log(1 - D(G(\mathbf{z}))].$$
\vspace{-1pt}
Radford et al. \cite{radford2015unsupervised} first specialized GANs to image synthesis by introducing Deep Convolutional GANs (DCGANs). DCGANs implement the generator and discriminator based on convolutional operations. Derivations of DCGANs such as Least Squares GAN (LSGAN) \cite{mao2016least} or Wasserstein GAN (WGAN) \cite{arjovsky2017wasserstein} tackle instabilities in the training procedure of early generative adversarial models. In our experiments, we found LSGAN to be most stable, in part leading to higher quality generated images than both DCGAN and WGAN.

The original GAN framework does not include an explicit means of performing inference. As such, we require extensions that allow mapping of a sample $\mathbf{x}$ drawn from the data prior to a latent representation via some encoder transformation $E(\mathbf{x})$. A common approach is to interpret the penultimate layer of the discriminator as a latent space. The activations of this layer serve as a source of representation vectors. These latent codes are thought to be meaningful because the discriminator must develop a strong internal representation of its input to succeed at its discrimination task. Furthermore, this method imposes no computational overhead compared to vanilla GANs. Figure \ref{fig:gan} shows a simplified schematic of our GAN architecture. More advanced encoders could tie this encoding to a corresponding noise vector.

\begin{figure}
  \centering
  \begin{tikzpicture}[thick]
    \tikzset{box/.style={
      draw, rectangle, rounded corners=0.8pt,
      text width=2.2cm, text height=0.75cm}}
    \tikzset{node/.style={draw, circle, inner sep=0.35cm}}

    % Model
    \path (0, 0) coordinate [draw, circle, inner sep=5pt]
          (z) node {$\mathbf{z}$};

    \path (2, 0) coordinate [box] (G) node {Generator};

    \path (4.3, -0.8) coordinate [node] (f) node {$\mathbf{x}$};
    \path (4.3, +0.8) coordinate [node] (r) node {$G(\mathbf{z})$};

    \path (9.5, 0) coordinate [draw, circle, inner sep=7pt]
          (D) node {$\hspace{1pt}\mathbf{D}$};

    % Anchor for real and fake arrows.
    \path (5, 0) coordinate (_);

    \foreach \x/\l in {0/Convolution,%
                       1/Convolution,%
                       2/Convolution,%
                       3/{\hspace{1mm}Dense(\small$\ell$)\textsuperscript{\color{Red}\textdagger}},%
                       4/{Dense({\small$1$})}%
                       } {
      \path ({5.8 + \x * 0.65}, 0)
            coordinate [box, text width=0.3cm, text height=2cm] (l\x)
            node [rotate=90] {\l};
    }

    % Footnote
    \draw (7, -1.5)
          node {\footnotesize{\color{Red}\textdagger} Representations};

    % Edges
    \draw (f) -- (_);
    \draw (r) -- (_);
    \draw [->] (_) -- (l0);
    \draw [->] (l4) -- (D);
    \draw [->] (z) -- (G);
    \draw [->] (G) edge[out=0] (f);
  \end{tikzpicture}
  \caption{A simplified diagram of the modified DCGAN \cite{radford2015unsupervised} architecture we employ for our experiments. On a high level, the discriminator is a sequence of convolutional layers, leading to a fully connected (dense) layer with $\ell$ neurons. We interpret the activations of these $\ell$ neurons as latent vectors and further feed them into a final dense layer with just one activation, forming the discriminator's output. }
  \label{fig:gan}
\end{figure}

@@

@@

\begin{figure}[b]
  \centering
  \begin{tikzpicture}
  	\draw (0, 0) node {\includegraphics[width=0.6\textwidth]{generated}};

@@

\section{Representation Learning for Morphological Profiling}\label{moa}
We test the quality of representations extracted via the discriminator by evaluating their ability to cluster treatments of similar function, here the mechanism-of-action. We obtain treatment profiles by averaging the extracted single cell representations found as intermediate activations of the discriminator. Further, we follow the experimental protocol of \cite{Ljosa2013} and report an average MOA classification accuracy of a leave-one-compound-out cross-validation using a one nearest neighbor classifier.
\begin{table}
  \centering
  \begin{tabular}{lllllll}
  	\toprule
    DCGAN & WGAN & LSGAN & VAE \cite{pawlowski2016msc} & CP \cite{Singh2014} & Deep Transfer \cite{pawlowski2016automating}  & Deep Transfer \cite{ando2017improving} \\
    \midrule
    51 \%& 56 \% & 68 \% & 49 \% & 90 \% & 91 \% & 96 \% \\
    \bottomrule \\
  \end{tabular}
  \caption{Comparison of the mechanism-of-action classification accuracy. CP refers to the results by Singh et al. \cite{Singh2014} using CellProfiler features. The Deep Transfer methods correspond to Deep Feature Transfer as proposed by Pawlowski et al. \cite{pawlowski2016automating} and refined by Ando et al. \cite{ando2017improving}.}
  \label{moa-results}
  \vspace{-0.5cm}
\end{table}

@@

This work covers only a small fraction of possible applications of GANs to the domain of microscopy images. For example, we hope that future work investigates BiGANs\cite{donahue2016adversarial} or other novel solutions to infer latent features with GANs. We believe that improved inference, combined with the interpretable nature of the GAN framework, may enable simulated experiments by performing algebra with vectors corresponding to cell lines, diseases or other perturbations. Finally, we note that deep learning generally flourishes with larger amounts of data. Thus, we conjecture that the performance of the outlined approaches will improve with larger datasets.

\section*{Acknowledgements}
We thank Allen Goodman, Jane Hung, Claire McQuin, Kyle Karhohs, Beth Cimini, Tim Becker, Minh Doan, Jeanelle Ackerman and Ray Jones of the Broad Institute for fruitful discussion, guidance and support with our experimental setup. We also extend our gratitude to Mike Ando and Google for providing computational resources to accelerate our research.

This work was supported in part by a grant from the US National Science Foundation (CAREER DBI 1148823 to AEC).

Nick Pawlowski is supported by Microsoft Research through its PhD Scholarship Program and the EPSRC Centre for Doctoral Training in High Performance Embedded and Distributed Systems (HiPEDS, Grant Reference EP/L016796/1).
